OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-04-07 paper/paper.Rmd
Better structure for split results

To compare to the reference clustering method used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously 
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
sequence's similarity to the single reference OTU.
In contrast, OptiFit accepts reference OTUs which each contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assign sequences to OTUs.
_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`% 
slower than VSEARCH.
In closed reference mode, VSEARCH was able to map `r frac_vsearch_vs_fit`% more
query sequences than OptiFit to the Greengenes reference database.
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of 
`r open_fit_gg_mcc` and `r open_vsearch_mcc` (respectively).
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and 
`r sec_vsearch_vs_open_fit_db`% on average respectively.
**TODO:** The stark difference in OTU definitions between mothur and VSEARCH resulted in...

@@

- we wanted ot use x% of data as ref, weren't sure what size to use and strategy for selecting.
- important result: it works!

When performing reference clustering against public databases, the database 
chosen greatly affects the quality of OTUs produced.
OTUs may be of poor quality when the reference database is too unrelated to the 
samples of interest, such as when samples contain new and rapidly evolving microbes.
For this reason, we were interested in whether we could use the samples themselves
as the reference to cluster against.

A split dataset strategy was employed to assess how well OptiFit performs for
tasks where new sequences are added to existing OTUs, such as when comparing
OTUs across studies or making predictions with machine learning models.
Datasets were randomly split into a reference fraction and a query fraction.
Reference sizes from 10% to 80% of the sequences were created, with the
remaining sequences used for the query.
Reference sequences were clustered _de novo_ with OptiClust, then query
sequences were fit to the _de novo_ OTUs with OptiFit.

**TODO; ref weight simple only here.**
OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split`% difference in median MCC).
OTU quality was remarkably stable across reference fraction sizes within
parameter sets (Fig. **XX**).
For example, splitting the human dataset 100 times yielded a coefficient of
variation of `r cv_fit_split_mcc_human` for the MCC score.
In terms of runtime, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
`r sec_open_fit_split_vs_clust` respectively.
The split dataset strategy performed `r sec_closed_fit_split_vs_db`% faster than
the database strategy in closed reference mode and `r sec_open_fit_split_vs_db`%
in open reference mode.
TODO: fraction mapped.

**TODO: pick a fraction (e.g. 50%). this part is less important. figure would be supplemental.**
We also tested three methods for selecting the fraction of sequences to be used
as the reference; a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset.
OTU quality was similar with the simple and abundance-weighted sampling (median
MCCs `r mcc_fit_split_simple` and `r mcc_fit_split_abun` respectively), but
worse for similarity-weighted sampling with a median MCC of `r mcc_fit_split_dist`.
In closed reference mode, the fraction of query sequences that can be fit to the
reference OTUs increases as the reference fraction increases; from 
`r frac_fit_split_0.1`% of query sequences fit when using 10% of the dataset as 
the reference, to `r frac_fit_split_0.8`% of query sequences fit when using 80% 
of the dataset as the reference.

>>>>

To compare to the reference clustering method used by QIIME2, we clustered each
dataset with VSEARCH against the Greengenes database of OTUs previously 
clustered at 97% sequence similarity.
Each reference OTU from the Greengenes 97% database contains one reference
sequence, and VSEARCH maps sequences to the reference based on each individual
sequence's similarity to the single reference OTU.
In contrast, OptiFit accepts reference OTUs which each contain multiple
sequences, and the sequence similarity between all query and reference sequences
is considered when assign sequences to OTUs.
_De novo_ clustering with OptiClust produced `r mcc_opticlust_vs_vsearch`%
higher quality OTUs than VSEARCH, but performed `r sec_opticlust_vs_vsearch`% 
slower than VSEARCH.
In closed reference mode, VSEARCH was able to map `r frac_vsearch_vs_fit`% more
query sequences than OptiFit to the Greengenes reference database.
When open reference clustering, OptiFit produced higher quality OTUs than
VSEARCH against the Greengenes database, with median MCC scores of 
`r open_fit_gg_mcc` and `r open_vsearch_mcc` (respectively).
In terms of run time, OptiFit outperformed VSEARCH in both closed and open
reference mode by `r sec_closed_fit_db_vs_vsearch`% and 
`r sec_vsearch_vs_open_fit_db`% on average respectively.
**TODO: conclude:** The stark difference in OTU definitions between mothur and VSEARCH resulted in...

@@

- we wanted to use x% of data as ref, weren't sure what size to use and strategy for selecting.
- important result: it works!

When performing reference clustering against public databases, the database 
chosen greatly affects the quality of OTUs produced.
OTUs quality may be poor when the reference database is too unrelated to the
samples of interest, such as when samples contain low abundant or rapidly
evolving microbes.
While _de novo_ clustering overcomes the quality limitations of reference
clustering to databases, OTU assignments are not stable when new sequences are
added.
Researchers may wish to fit new sequences to existing OTUs when comparing OTUs
across studies or when making predictions with machine learning models.
To determine how well OptiFit performs for fitting new sequences to existing
OTUs, we employed a split dataset strategy, where each dataset was randomly
split into a reference fraction and a query fraction.
Reference sequences were clustered _de novo_ with OptiClust, then query
sequences were fit to the _de novo_ OTUs with OptiFit.

**TODO; ref weight simple only here. First we wanted to know; does this strategy perform as well as de novo with 50% split?**
OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (`r mcc_opticlust_vs_fit_split_simple`% difference in median MCC).
OTU quality was remarkably stable across reference fraction sizes within
parameter sets.
For example, splitting the human dataset 100 times yielded a coefficient of
variation of `r cv_fit_split_mcc_human` for the MCC score.
In terms of runtime, closed and open reference OptiFit performed faster than
OptiClust on whole datasets by `r sec_closed_fit_split_vs_clust`% and 
`r sec_open_fit_split_vs_clust` respectively.
The split dataset strategy performed `r sec_closed_fit_split_vs_db`% faster than
the database strategy in closed reference mode and `r sec_open_fit_split_vs_db`%
faster in open reference mode.
**TODO: fraction mapped.**

**Then we wanted to know; what fraction of sequences should be in the reference?**
To test the best reference size, reference sizes from 10% to 80% of the sequences were created, with the remaining sequences used for the query.

**Finally, we wanted to know the best way to select the reference sequences.**
**TODO: pick a fraction (e.g. 50%). this part is less important. figure would be supplemental.**
We also tested three methods for selecting the fraction of sequences to be used
as the reference; a simple random sample, weighting sequences by relative
abundance, and weighting by similarity to other sequences in the dataset.
OTU quality was similar with the simple and abundance-weighted sampling (median
MCCs `r mcc_fit_split_simple` and `r mcc_fit_split_abun` respectively), but
worse for similarity-weighted sampling with a median MCC of `r mcc_fit_split_dist`.
In closed reference mode, the fraction of query sequences that can be fit to the
reference OTUs increases as the reference fraction increases; from 
`r frac_fit_split_0.1`% of query sequences fit when using 10% of the dataset as 
the reference, to `r frac_fit_split_0.8`% of query sequences fit when using 80% 
of the dataset as the reference.
